Comparing MCMC algorithms in Stochastic Volatility Models using Simulation Based Calibration

### 2023-10-19 manuscript/main.tex
Syntax edits

    The key findings of this research are HMC produces more calibrated and efficient posterior estimates relative to KSC's MCMC algorithm on the approximate off-set mixture model. Additionally, MCMC algorithms may perform better or worse based on calibration and efficiency depending on the parameterisation of the model. HMC applied on a re-parameterised SV model produces the most calibrated and efficient posterior estimates across all SBC experiments.

@@

\subsection{Stochastic Volatility}
    The model of interest is the discrete time, univariate stochastic volatility model estimated by \citet{kim1998stochastic} using a Bayesian framework. SV is expressed as a state space model which links the observed (logarithmic) return of a given asset over period $t$, denoted $y_t$, to its latent conditional variance, $h_{t+1}$. This model is given by the following expressions
    \begin{align}
    y_t =& \space \,\exp(h_t/2) \varepsilon_t \\
    h_{t+1} =& \space \, \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t  \\
    h_1 \sim& \space \,\mathrm{N}\left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right) \\
    \varepsilon_t \sim& \space \,\mathrm{N}(0,1) \\
    \eta_t \sim& \space \,\mathrm{N}(0,1).
    \end{align}
    The measurement equation (1) describes the conditional distribution of $y_t$ given $h_t$, where $h_t$ is the conditional log variance of $y_t$. $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the auto-regressive parameter $\phi$ such that $|\phi|<1$. This auto-regressive parameter represents the persistence or stickiness of log volatility and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

@@

\subsection{Research Goal}
    The objective of this research is to design a simulation study to evaluate the calibration of the HMC and KSC offset-mixture algorithms used to estimate SV models. As discussed, there are limitations to evaluating MCMC algorithms based on fits to real data and single simulations. To check the calibration of an algorithm, repeated independent simulations are required. This methodology is discussed in the next section.

@@

            \item Chi squared statistics calculated over the $K$ rank statistics to summarise the shape of the distribution and deviation from uniformity.

@@

            \subsubsection{Chi squared statistics}
            One difficulty of relying on the visualisation of rank statistic distributions for SV models is there are a very large number of unknowns in the model to consider. An alternative to visually checking for uniformity of a given parameter is to calculate a Chi-squared statistic based on the counts in each histogram bin. Suppose the histogram has $J$ bins, where for each $j^{th}$ bin, $b_j$ is the number of counts and $e_j$ the expected count in bin $j$. The expected count is the ratio of $K$ rank statistics and $J$ bins, $e_j=\frac{K}{J}$. The corresponding Chi-squared statistic is denoted by
            \begin{align}
            \chi^2 = \sum_{j=1}^J \frac{(b_{j} - e_{j})^2}{e_j}.
            \end{align}

            A discrete empirical distribution with the same number of realisations in each of $J$ bins will return a chi squared statistic of zero, since the number of realisations in each bin is equal to the expected count under the relevant  uniform distribution. The distribution of chi squared statistics is visualised to compare results across multiple simulations. Thus, the distribution of chi-squared statistics produced from the histograms of SBC rank statistics for a large number of univariate model parameters will give a high level summary of the overall model calibration, and hence for the overall performance of the algorithm. By comparing the distributions of Chi-squared statistics produced by competing algorithms, one algorithm's degree of calibration, relative to that of the other algorithm, may be assessed.

@@

\section{MCMC Strategies}
This section defines and compares the two MCMC algorithms used to estimate the SV model. Section 4.1 and Section 4.2 describes Stan's No-U-Turn sampler and KSC's mixture offset MCMC respectively. Section 4.3 compares the similarities and differences between both methods. Section 4.4 describes model reparameterisation in the context of each MCMC algorithm. 
    \subsection{Sampling method 1: Stan's Hamiltonian Monte Carlo with No-U-Turn Sampler}
        Hamiltonian Monte Carlo (HMC) is a general MCMC algorithm used to efficiently sample posterior draws associated with high-dimensional and complex stochastic models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo, was developed in the physics literature \citep{duane1987hybrid} before being applied in the statistics literature by Radford Neal through his works in Bayesian Neural Networks \citep{neal1995bayesian} and statistical computing \citep{neal2011mcmc}. Versions of this algorithm have since become widely available through open source development projects, including Stan \citep{stan} and PyMC \citep{pymc2023}.

@@

    \subsection{Sampling method 2: Off-set Mixture MCMC}
        \citet{kim1998stochastic} sample the conditional posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings \citep{metropolis1953equation, hastings1970monte} within Gibbs \citep{geman1984stochastic} algorithms and an implementation of the Kalman Filter \citep{kalman1960new} and smoother used to sample from the conditional posterior distribution of the latent states \citep{dejong1995}.\footnote{In this research the exact software to apply the simulation smoother is unavailable. So a more recent simulation smoother is used from the statsmodels Python package by \citet{seabold2010statsmodels}. The simulation smoother software within the statsmodels package is based on \citet{durbin2012time}. Koopman who co-authored \citet{durbin2012time} also co-authored the software used by \citet{kim1998stochastic} which can be found at \citet{koopman1996ssfpack}.} Note that for the simulation smoother to produce an appropriate posterior sample draw from the latent state vector, the state and measurement equations are required to be linear and conditionally Gaussian. Since the relationship between $y_t$ and $h_t$ in the measurement equation is not linear, a transformation is applied by squaring, adding an offset, and taking the log of $y_t$, resulting in
        \begin{align}
        y_t^{*} &= log(y_t^2 + c) = h_t + z_t, \label{eqn:transform}
        \end{align}
        where $z_t = log(\epsilon_t^2)$ has a distribution that is like that of the logarithm of a chi-squared random variable. It is known that the error $z_t$ has mean -1.2704 and variance 4.93. The offset $c$ is set to $c=0.001$ to ensure the model is robust to small $y_t^2$ values \citep{fuller1996introduction}. By transforming the entire measurement equation, as in (\ref{eqn:transform}), the relationship between $y_t^{*}$ and $h_t$ is now linear; however, the error is not Gaussian. Since it is not simple to sample from this parameterisation of the model, KSC use a mixture of seven Gaussian distributions to approximate the distribution of $z_t$, with the mixture selected to match the first 4 moments of the log chi squared distribution. The mixture distribution is defined by
        \begin{align}
        f(z_t) = \sum_{i=1}^{I} q_if_N(z_t|m_i-1.2704, \nu_i^2),
        \end{align}
        where $K=7$, $f_N(z_t|m_i-1.2704, \nu_i^2)$ denotes the density of the normal distribution with mean $m_i-1.2704$ and variance $v_i^2$, and  component probabilities $q_i$, for $i=1,2,\ldots, I=7$.  The values of these hyperparameters and component weights are applied in this research and reported in Appendix A.

@@

        \item Sample from a standard normal distribution and multiply by the variance of the log volatility. The state vector is sampled with mean centered on zero and variance equal to $\sigma_{\eta}^2$ as given by
        \begin{align}
        h_{std} \sim&\space \,\mathrm{N}(0,1) \\
        h =& \space \, h_{std} \times \sigma_{\eta}.
        \end{align}
        \item Apply the appropriate re-scaling to get samples from the log variance
        \begin{align}
        h_1 =& \space \,\frac{h_{std, 1}\times \sigma_{\eta}} {\sqrt{1 - \phi^2}} + \mu \\
        h_{t+1} =& \space \,h_{std, t+1}\times \sigma_{\eta} + \mu  + \phi(h_{t} - \mu),\space \, t\neq 1.
        \end{align}
        \end{enumerate}
        This returns the log variance as desired.

        \subsubsection{Non-centered Gaussian mixture off-set model}
        The re-parameterised model for the Gaussian mixture off-set is expressed differently due to the Kalman Filter. Re-parameterised HMC samples from the joint posterior directly with state vectors centered at zero and transformed to return the correct log volatility estimates. Applying the Kalman Filter requires the state equation to be a function of the state variable with the average log volatility parameter $\mu$ entering the measurement equation.

        The following parameterisation is described as non centered in location. This follows the methodology outlined in \citet{strickland2008parameterisation}. Starting with the log chi squared model
        \begin{align}
        y^{\ast}_t =& \space \, h_t + z_t \\
        h_{t+1} =& \space \, \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t.
        \end{align}
        Reparameterise the state variable $h_t$ by subtracting the average log variance $\mu$. Let $g_t = \space \, h_t - \mu$. This leads to the state equation
        \begin{align}
        g_{t+1} =& \space \, \phi g_t + \sigma_{\eta}\eta_{t}.
        \end{align}
        Return average volatility into measurement equation and rewrite as a function of the non centered state equation
        \begin{align}
        y^{\ast}_t =& \space \, g_t + \mu + z_t \\
        g_{t+1} =& \space \, \phi g_t + \sigma_{\eta}\eta_{t},
        \end{align}
        where the initial state now is drawn from 
        \begin{align}
            g_1 \sim& \space \,\mathrm{N}\left(0, \frac{\sigma_{\eta}^2}{1-\phi^2}\right).
        \end{align}
        The mean of the log volatility is now inside the measurement equation and demeaned from the state equation.

    \subsubsection{Prior predictive check}
        % Different parameteristions express the same mathematical models in ways that make it easier or harder for a given algorithm to sample. 
        A prior predictive simulation of the second state parameter is performed for both a centered and re-parameterised model using the approach described in Section 4.4.1. This is to demonstrate that different parameterisations express the same underlying mathematical model. This is similar to the first two steps of SBC, except instead of generating a data set, it generates the state variable implied by the prior draw. Let $\boldsymbol{\theta}$ be a vector of static parameters. Simulate a draw from the joint prior
        \begin{align}
        \boldsymbol{\theta}^{sim} \sim \pi(\boldsymbol{\theta}).
        \end{align}

@@

        \begin{figure}[h]
            \centering
            \includegraphics[scale=0.1]{figures/ppc_h2.png}
            \caption{\textbf{Prior predictive simulation of second state variable from both centered and re-parameterised stochastic volatility model}. Both models are mathematically equivalent and generate samples from the same data generating process.}
            \label{fig:priorpred}
        \end{figure}

@@

    Rank statistic histograms are constructed with 20 bins for the static parameters and the 1st, 500th and 1000th latent state variables. The black horizontal line represents the expected counts in each bin if the algorithm is returning calibrated estimates and can be viewed as the uniform distribution. This is calculated to be 250 which is the ratio of  5000 SBC iterations and 20 bins. The Chi-squared statistic is used to summarise the deviations from uniformity for all state variables since it is unrealistic to visually inspect all the histograms. The SBC algorithm is summarised in Algorithm \ref{alg:sbc}.

@@

    Figure \ref{fig:ncphmc5k} show the SBC results for the reparameterised model using HMC. The re-parameterised model produces approximately uniform rank statistics across all reported parameters. The distributions of $\sigma^2$ and $\phi$ now exhibit a more uniform shape, implying greater calibration and correct posterior estimates returned on average. These results suggest that HMC struggles to sample some parameters from the centered parameterisation of the stochastic volatility model. The re-parameterised model possess stronger evidence of calibration for the parameters examined, producing more reliable and consistent inference.
        \begin{figure}[H]
        \centering
        \includegraphics[scale=0.09]{results/hmc_cp_5k.png}
        \caption{\textbf{Distribution of rank statistics with 5000 SBC iterations for centered model using Hamiltonian Monte Carlo}. $\mu$ and the state parameters look approximately uniform. $\sigma^2$ still has inflated frequencies on both the left and right hand side indicating under-dispersion relative to the prior distribution. $\phi$ exhibits a similar U-shape to $\sigma^2$ although the U-shape is much smaller.}
        \label{fig:cphmc5k}
    \end{figure}
    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.09]{results/hmc_ncp_5k.png}
        \caption{\textbf{Distribution of rank statistics with 5000 SBC iterations for re-parameterised model using Hamiltonian Monte Carlo}. The rank statistic distributions for $\sigma^2$ and $\phi$ appear to follow a uniform distribution. This suggests that the HMC algorithm is returning the correct posterior estimates on average for the displayed parameters.}
        \label{fig:ncphmc5k}
    \end{figure}

    This computational difficulty in the centered parameterisation is reflected in the effective sample sizes. Figure \ref{fig:hmcess} and Table \ref{tab:hmcess} summarises the distribution of ESS for the static parameters after 5000 SBC iterations. The reparameterised model generates independent samples much more efficiently than the centered model, reflected in the larger ESS values across every summary statistic.  This is consistent with HMC struggling to sample overall from the posteriors of $\sigma^2$ and $\phi$ in the centered model.

@@

    \begin{table}[H]
        \centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|} \hline 
        Parameter &  Type&Min& q25&  Median& Mean & q75&Max\\ \hline 
        $\mu$&  Centered&3.79 & 99.7 & 184. & 305. & 457. & 1997.  \\
     $\mu$&  Reparam&14.5 & 738. & 1099. & 1143. & 1500. & 2993.  \\\hline 
     $\phi$&  Centered&1.44 & 11.0 & 21.7 & 26.3 & 34.1 & 635.  \\
     $\phi$&  Reparam&48.8 & 355. & 476. & 521. & 651. & 1445.   \\ \hline 
     $\sigma^2$&  Centered& 1.17 & 2.99 & 7.04 & 8.05 & 11.3 & 90.7 \\ 
     $\sigma^2$&  Reparam&46.1 & 403. & 553. & 585. & 734. & 1486. \\ \hline
        \end{tabular}
        \caption{\textbf{ESS for centered and re-parameterised SV model across 5000 SBC iterations for HMC}. The reparameterised model has higher ESS counts for each summary statistic relative to the centered parameterisation.}
        \label{tab:hmcess}
    \end{table}

@@

    The ESS estimates for this model indicate a high degree of auto-correlation and difficulty in generating independent samples from both parameterisations. The numerical results are summarised in Figure \ref{fig:kscess} and Table \ref{tab:kscess}. The estimate for $\mu$ in the re-parameterised model exhibits a long right tail (longer than the 9,999 post burn-in posterior draws suggesting the presence of negatively auto-correlated draws). There is some improvement in efficiency from the re-parameterised model with higher ESS values across most of the static parameter quantiles. Overall, the off-set mixture MCMC is inefficient at sampling the static parameters. Note that we cannot directly compare the effect of the reparameterisation on the ESS between Figure \ref{fig:kscess} and Figure \ref{fig:hmcess} due to the difference in post burn-in posterior draws.

@@

    \subsubsection{Re-weighting rank statistics of the Gaussian off-set mixture model}
    Let $v(\theta, h)$ be the log weights defined as the log difference between the posterior densities of the true model with log chi squared errors $log\: c(\theta, h | y^{\ast})$ and the mixture off-set model $log\:  d(\theta, h | y^{\ast})$. The is given in the following expression
    \begin{align}
        v(\theta, h) = log\: c(\theta, h | y^{\ast}) - log\:  d(\theta, h | y^{\ast}) = \text{const} + log\: c(y|h) - log\: d(y^{\ast} | h).
    \end{align}

@@

    The results from applying this re-weighting step to the rank statistics of the centered model are given on Figure \ref{fig:reweight5k}. There are no improvements in the distribution of rank statistics. The shape of the histograms is consistent with the shape of the unweighted rank statistics. Overall re-weighting the posterior samples from the approximate model does not improve the calibration results. The re-weighted rank statistics for the re-parameterised model also did not improve and can be found in Appendix E.

@@

    \subsection{Evaluating SBC for state variables}
    It is not practical to inspect the histograms of all parameters and states in the stochastic volatility model. Instead, the Chi-squared statistic is used to summarise the shape of the rank statistic distribution. This will be used to summarise the state variables only since the static parameters can be compared across simulations individually.

    The Chi-squared statistics for centered and re-parameterised HMC, centered KSC and centered importance weighted KSC are reported as histograms in Figure \ref{fig:allchisq} using results from 5000 SBC iterations. Re-parameterised KSC is omitted as its poor SBC performance skews the scale of the plot which makes it difficult to analyse results. A chi squared statistic of 0 means the rank statistics for a given parameter is exactly uniformly distributed and any value away from 0 captures deviations from uniformity. The chi squared statistics for the latent state variable ranks are expected to be some non-zero value since they are calculated from finite samples and subject to sampling variation even if the draws come from a uniform distribution. The purpose of this plot is to provide a high level summary of the latent state variable calibration for each simulation - how close their rank statistics are to being uniformly distributed and their distance relative to the other algorithms and parameterisations.

    The HMC results are relatively close to zero. Both KSC and the importance weighted ranks on the other hand are far from zero suggesting larger deviations from uniformity. There is a wide gap between the HMC and KSC chi squared statistics implying the distribution of latent state ranks are much more  uniform for HMC relative to KSC. Combined with the results in the previous section, there is a strong evidence that the HMC algorithm outperforms the KSC algorithm where it comes to calibration across all parameters and latent state variables.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{results/dist_chisq_all.png}
        \caption{\textbf{Distribution of chi squared statistics for latent state variables from 5000 SBC iterations}. Values closer to 0 are more consistent with a discrete uniform distribution. Both HMC simulations are much closer to zero than KSC. This suggests that the HMC algorithm overall produces more calibrated posterior estimates for the latent state variables.}
        \label{fig:allchisq}
    \end{figure}

@@

\section{Appendix C: Hamiltonian Monte Carlo algorithm description}
Let $\theta$ be a target parameter and $\alpha$ be the auxiliary momentum variable. The Hamiltonian is defined as:

>>>>

    The key findings of this research are HMC produces more calibrated and efficient posterior estimates relative to KSC's MCMC algorithm on the approximate off-set mixture model. Additionally, MCMC algorithms may perform better or worse based on calibration and efficiency depending on the parameterisation of the model. HMC applied on a reparameterised SV model produces the most calibrated and efficient posterior estimates across all SBC experiments.

@@

\subsection{Stochastic Volatility}
    The model of interest is the discrete time, univariate stochastic volatility model estimated by \citet{kim1998stochastic} using a Bayesian framework. SV is expressed as a state space model which links the observed (logarithmic) return of a given asset over period $t$, denoted $y_t$, to its latent conditional variance, $h_{t+1}$. This model is given by the following expressions
    \begin{align}
    y_t =& \space \,\exp(h_t/2) \varepsilon_t \\
    h_{t+1} =& \space \, \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t  \\
    h_1 \sim& \space \,\mathrm{N}\left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right) \\
    \varepsilon_t \sim& \space \,\mathrm{N}(0,1) \\
    \eta_t \sim& \space \,\mathrm{N}(0,1).
    \end{align}
    The measurement equation (1) describes the conditional distribution of $y_t$ given $h_t$, where $h_t$ is the conditional log variance of $y_t$. $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the auto-regressive parameter $\phi$ such that $|\phi|<1$. This auto-regressive parameter represents the persistence or stickiness of the log variance and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

@@

\subsection{Research Goal}
    The objective of this research is to design a simulation study to evaluate the calibration of the HMC and KSC off-set mixture algorithms used to estimate SV models. As discussed, there are limitations to evaluating MCMC algorithms based on fits to real data and single simulations. To check the calibration of an algorithm, repeated independent simulations are required. This methodology is discussed in the next section.

@@

            \item Chi-squared statistics calculated over the $K$ rank statistics to summarise the shape of the distribution and deviation from uniformity.

@@

            \subsubsection{Chi-squared statistics}
            One difficulty of relying on the visualisation of rank statistic distributions for SV models is there are a very large number of unknowns in the model to consider. An alternative to visually checking for uniformity of a given parameter is to calculate a chi-squared statistic based on the counts in each histogram bin. Suppose the histogram has $J$ bins, where for each $j^{th}$ bin, $b_j$ is the number of counts and $e_j$ the expected count in bin $j$. The expected count is the ratio of $K$ rank statistics and $J$ bins, $e_j=\frac{K}{J}$. The corresponding chi-squared statistic is denoted by
            \begin{align}
            \chi^2 = \sum_{j=1}^J \frac{(b_{j} - e_{j})^2}{e_j}.
            \end{align}

            A discrete empirical distribution with the same number of realisations in each of $J$ bins will return a chi-squared statistic of zero, since the number of realisations in each bin is equal to the expected count under the relevant  uniform distribution. The distribution of chi-squared statistics is visualised to compare results across multiple simulations. Thus, the distribution of chi-squared statistics produced from the histograms of SBC rank statistics for a large number of univariate model parameters will give a high level summary of the overall model calibration, and hence for the overall performance of the algorithm. By comparing the distributions of chi-squared statistics produced by competing algorithms, one algorithm's degree of calibration, relative to that of the other algorithm, may be assessed.

@@

\section{MCMC Strategies}
This section defines and compares the two MCMC algorithms used to estimate the SV model. Section 4.1 and Section 4.2 describes Stan's No-U-Turn sampler and KSC's off-set mixture MCMC respectively. Section 4.3 compares the similarities and differences between both methods. Section 4.4 describes model reparameterisation in the context of each MCMC algorithm. 
    \subsection{Sampling method 1: Stan's Hamiltonian Monte Carlo with No-U-Turn Sampler}
        Hamiltonian Monte Carlo (HMC) is a general MCMC algorithm used to efficiently sample posterior draws associated with high-dimensional and complex stochastic models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo, was developed in the physics literature \citep{duane1987hybrid} before being applied in the statistics literature by Radford Neal through his works in Bayesian Neural Networks \citep{neal1995bayesian} and statistical computing \citep{neal2011mcmc}. Versions of this algorithm have since become widely available through open source development projects, including Stan \citep{stan} and PyMC \citep{pymc2023}.

@@

    \subsection{Sampling method 2: Off-set Mixture MCMC}
        \citet{kim1998stochastic} sample the conditional posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings \citep{metropolis1953equation, hastings1970monte} within Gibbs \citep{geman1984stochastic} algorithms and an implementation of the Kalman Filter \citep{kalman1960new} and smoother used to sample from the conditional posterior distribution of the latent states \citep{dejong1995}.\footnote{In this research the exact software to apply the simulation smoother is unavailable. So a more recent simulation smoother is used from the statsmodels Python package by \citet{seabold2010statsmodels}. The simulation smoother software within the statsmodels package is based on \citet{durbin2012time}. Koopman who co-authored \citet{durbin2012time} also co-authored the software used by \citet{kim1998stochastic} which can be found at \citet{koopman1996ssfpack}.} Note that for the simulation smoother to produce an appropriate posterior sample draw from the latent state vector, the state and measurement equations are required to be linear and conditionally Gaussian. Since the relationship between $y_t$ and $h_t$ in the measurement equation is not linear, a transformation is applied by squaring, adding an off-set, and taking the log of $y_t$, resulting in
        \begin{align}
        y_t^{*} &= \log(y_t^2 + c) = h_t + z_t, \label{eqn:transform}
        \end{align}
        where $z_t = \log(\epsilon_t^2)$ has a distribution that is like that of the logarithm of a chi-squared random variable. It is known that the error $z_t$ has mean -1.2704 and variance 4.93. The off-set $c$ is set to $c=0.001$ to ensure the model is robust to small $y_t^2$ values \citep{fuller1996introduction}. By transforming the entire measurement equation, as in (\ref{eqn:transform}), the relationship between $y_t^{*}$ and $h_t$ is now linear; however, the error is not Gaussian. Since it is not simple to sample from this parameterisation of the model, KSC use a mixture of seven Gaussian distributions to approximate the distribution of $z_t$, with the mixture selected to match the first 4 moments of the log chi-squared distribution. The mixture distribution is defined by
        \begin{align}
        f(z_t) = \sum_{i=1}^{I} q_if_N(z_t|m_i-1.2704, \nu_i^2),
        \end{align}
        where $K=7$, $f_N(z_t|m_i-1.2704, \nu_i^2)$ denotes the density of the normal distribution with mean $m_i-1.2704$ and variance $v_i^2$, and  component probabilities $q_i$, for $i=1,2,\ldots, I=7$.  The values of these hyperparameters and component weights are applied in this research and reported in Appendix A.

@@

        \item Sample from a standard normal distribution and multiply by the variance of the states. The state vector is sampled with mean centered on zero and variance equal to $\sigma_{\eta}^2$ as given by
        \begin{align}
        h_{std} \sim&\space \,\mathrm{N}(0,1) \\
        h =& \space \, h_{std} \times \sigma_{\eta}.
        \end{align}
        \item Apply the appropriate re-scaling to get samples from the log variance
        \begin{align}
        h_1 =& \space \,\frac{h_{std, 1}\times \sigma_{\eta}} {\sqrt{1 - \phi^2}} + \mu \\
        h_{t+1} =& \space \,h_{std, t+1}\times \sigma_{\eta} + \mu  + \phi(h_{t} - \mu),\space \, t\neq 1.
        \end{align}
        \end{enumerate}
        This returns the log variance as desired.

        \subsubsection{Non-centered Gaussian mixture off-set model}
        The reparameterised model for the Gaussian mixture off-set is expressed differently due to the Kalman Filter. Reparameterised HMC samples from the joint posterior directly with state vectors centered at zero and transformed to return the correct log volatility estimates. Applying the Kalman Filter requires the state equation to be a function of the state variable with the average log volatility parameter $\mu$ entering the measurement equation.

        The following parameterisation is described as non centered in location. This follows the methodology outlined in \citet{strickland2008parameterisation}. Starting with the log chi-squared model
        \begin{align}
        y^{\ast}_t =& \space \, h_t + z_t \\
        h_{t+1} =& \space \, \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t.
        \end{align}
        Reparameterise the state variable $h_t$ by subtracting the average log variance $\mu$. Let $g_t = \space \, h_t - \mu$. This leads to the state equation
        \begin{align}
        g_{t+1} =& \space \, \phi g_t + \sigma_{\eta}\eta_{t}.
        \end{align}
        Return average volatility into measurement equation and rewrite as a function of the non centered state equation
        \begin{align}
        y^{\ast}_t =& \space \, g_t + \mu + z_t \\
        g_{t+1} =& \space \, \phi g_t + \sigma_{\eta}\eta_{t},
        \end{align}
        where the initial state now is drawn from 
        \begin{align}
            g_1 \sim& \space \,\mathrm{N}\left(0, \frac{\sigma_{\eta}^2}{1-\phi^2}\right).
        \end{align}
        The mean of the log volatility is now inside the measurement equation and demeaned from the state equation.

    \subsubsection{Prior predictive check}
        % Different parameteristions express the same mathematical models in ways that make it easier or harder for a given algorithm to sample. 
        A prior predictive simulation of the second state parameter is performed for both a centered and reparameterised model using the approach described in Section 4.4.1. This is to demonstrate that different parameterisations express the same underlying mathematical model. This is similar to the first two steps of SBC, except instead of generating a data set, it generates the state variable implied by the prior draw. Let $\boldsymbol{\theta}$ be a vector of static parameters. Simulate a draw from the joint prior
        \begin{align}
        \boldsymbol{\theta}^{sim} \sim \pi(\boldsymbol{\theta}).
        \end{align}

@@

        \begin{figure}[h]
            \centering
            \includegraphics[scale=0.1]{figures/ppc_h2.png}
            \caption{\textbf{Prior predictive simulation of second state variable from both centered and reparameterised stochastic volatility model}. Both models are mathematically equivalent and generate samples from the same data generating process.}
            \label{fig:priorpred}
        \end{figure}

@@

    Rank statistic histograms are constructed with 20 bins for the static parameters and the 1st, 500th and 1000th latent state variables. The black horizontal line represents the expected counts in each bin if the algorithm is returning calibrated estimates and can be viewed as the uniform distribution. This is calculated to be 250 which is the ratio of  5000 SBC iterations and 20 bins. The chi-squared statistic is used to summarise the deviations from uniformity for all state variables since it is unrealistic to visually inspect all the histograms. The SBC algorithm is summarised in Algorithm \ref{alg:sbc}.

@@

    Figure \ref{fig:ncphmc5k} show the SBC results for the reparameterised model using HMC. The reparameterised model produces approximately uniform rank statistics across all reported parameters. The distributions of $\sigma^2$ and $\phi$ now exhibit a more uniform shape, implying greater calibration and correct posterior estimates returned on average. These results suggest that HMC struggles to sample some parameters from the centered parameterisation of the stochastic volatility model. The reparameterised model possess stronger evidence of calibration for the parameters examined, producing more reliable and consistent inference.
        \begin{figure}[H]
        \centering
        \includegraphics[scale=0.09]{results/hmc_cp_5k.png}
        \caption{\textbf{Distribution of rank statistics with 5000 SBC iterations for centered model using Hamiltonian Monte Carlo}. $\mu$ and the state parameters look approximately uniform. $\sigma^2$ still has inflated frequencies on both the left and right hand side indicating under-dispersion relative to the prior distribution. $\phi$ exhibits a similar U-shape to $\sigma^2$ although the U-shape is much smaller.}
        \label{fig:cphmc5k}
    \end{figure}
    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.09]{results/hmc_ncp_5k.png}
        \caption{\textbf{Distribution of rank statistics with 5000 SBC iterations for reparameterised model using Hamiltonian Monte Carlo}. The rank statistic distributions for $\sigma^2$ and $\phi$ appear to follow a uniform distribution. This suggests that the HMC algorithm is returning the correct posterior estimates on average for the displayed parameters.}
        \label{fig:ncphmc5k}
    \end{figure}

    Computational difficulty in the centered parameterisation is reflected in the effective sample sizes. Figure \ref{fig:hmcess} and Table \ref{tab:hmcess} summarise the ESS distributions for the static parameters in both parameterisations after 5000 SBC iterations. The reparameterised model generates independent samples much more efficiently than the centered model, reflected in the larger ESS values across every summary statistic. The median ESS for $\mu$, $\phi$ and $\sigma^2$ in the centered parameterisation is much smaller as a proportion of the posterior draws. This is consistent with HMC struggling to sample from the posteriors of the static parameters overall in the centered model.

@@

    \begin{table}[H]
        \centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|} \hline 
        Parameter &  Type&Min& q25&  Median& Mean & q75&Max\\ \hline 
        $\mu$&  Centered&3.79 & 99.7 & 184. & 305. & 457. & 1997.  \\
     $\mu$&  Reparam&14.5 & 738. & 1099. & 1143. & 1500. & 2993.  \\\hline 
     $\phi$&  Centered&1.44 & 11.0 & 21.7 & 26.3 & 34.1 & 635.  \\
     $\phi$&  Reparam&48.8 & 355. & 476. & 521. & 651. & 1445.   \\ \hline 
     $\sigma^2$&  Centered& 1.17 & 2.99 & 7.04 & 8.05 & 11.3 & 90.7 \\ 
     $\sigma^2$&  Reparam&46.1 & 403. & 553. & 585. & 734. & 1486. \\ \hline
        \end{tabular}
        \caption{\textbf{ESS for centered and reparameterised SV model across 5000 SBC iterations for HMC}. The reparameterised model has higher ESS counts for each summary statistic relative to the centered parameterisation.}
        \label{tab:hmcess}
    \end{table}

@@

    The ESS estimates for this model indicate a high degree of auto-correlation and difficulty in generating independent samples from both parameterisations. The numerical results are summarised in Figure \ref{fig:kscess} and Table \ref{tab:kscess}. The estimate for $\mu$ in the reparameterised model exhibits a long right tail, with values larger than the 9,999 post burn-in posterior draws suggesting the presence of negatively auto-correlated draws. The median ESS for each static parameter relative to the number of posterior draws is very small. There is some improvement from the reparameterised model with higher ESS values across some of the static parameter quantiles, but these improvements are negligible. Overall, the off-set mixture MCMC is inefficient at sampling the static parameters. Note, Figure \ref{fig:hmcess} and Figure \ref{fig:kscess} are not directly compared due to scale difference from varying post burn-in posterior draws.

    % that we cannot directly compare the effect of the reparameterisation on the ESS between Figure \ref{fig:kscess} and Figure \ref{fig:hmcess} due to the difference in post burn-in posterior draws.

@@

    \subsubsection{Re-weighting rank statistics of the Gaussian off-set mixture model}
    Let $v(\theta, h)$ be the log weights defined as the log difference between the posterior densities of the true model with log chi-squared errors $\log c(\theta, h | y^{\ast})$ and the mixture off-set model $\log d(\theta, h | y^{\ast})$. The is given in the following expression
    \begin{align}
        v(\theta, h) = \log c(\theta, h | y^{\ast}) - \log  d(\theta, h | y^{\ast}) = \text{const} + \log c(y|h) - \log d(y^{\ast} | h).
    \end{align}

@@

    The results from applying this re-weighting step to the rank statistics of the centered model are given on Figure \ref{fig:reweight5k}. There are no improvements in the distribution of rank statistics. The shape of the histograms is consistent with the shape of the unweighted rank statistics. Overall re-weighting the posterior samples from the approximate model does not improve the calibration results. The re-weighted rank statistics for the reparameterised model also did not improve and can be found in Appendix E.

@@

    \subsection{Evaluating SBC for state variables}
    It is not practical to inspect the histograms of all parameters and states in the stochastic volatility model. Instead, the chi-squared statistic is used to summarise the shape of the rank statistic distribution. This will be used to summarise the state variables only since the static parameters can be compared across simulations individually.

    The chi-squared statistics for centered and reparameterised HMC, centered KSC and centered importance weighted KSC are reported as histograms in Figure \ref{fig:allchisq} using results from 5000 SBC iterations. Reparameterised KSC is omitted as its poor SBC performance skews the scale of the plot which makes it difficult to analyse results. A chi-squared statistic of 0 means the rank statistics for a given parameter is exactly uniformly distributed and any value away from 0 captures deviations from uniformity. The chi-squared statistics for the latent state variable ranks are expected to be some non-zero value since they are calculated from finite samples and subject to sampling variation even if the draws come from a uniform distribution. The purpose of this plot is to provide a high level summary of the latent state variable calibration for each simulation - how close their rank statistics are to being uniformly distributed and their distance relative to the other algorithms and parameterisations.

    The HMC results are relatively close to zero. Both KSC and the importance weighted ranks on the other hand are far from zero suggesting larger deviations from uniformity. There is a wide gap between the HMC and KSC chi-squared statistics implying the distribution of latent state ranks are much more  uniform for HMC relative to KSC. Combined with the results in the previous section, there is a strong evidence that the HMC algorithm outperforms the KSC algorithm where it comes to calibration across all parameters and latent state variables.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{results/dist_chisq_all.png}
        \caption{\textbf{Distribution of chi-squared statistics for latent state variables from 5000 SBC iterations}. Values closer to 0 are more consistent with a discrete uniform distribution. Both HMC simulations are much closer to zero than KSC. This suggests that the HMC algorithm overall produces more calibrated posterior estimates for the latent state variables.}
        \label{fig:allchisq}
    \end{figure}

@@

\section{Appendix C: Hamiltonian Monte Carlo algorithm description}
Let $\theta$ be a target parameter and $\alpha$ be the auxiliary momentum variable. The Hamiltonian is defined as
\begin{align}
H(\theta, \alpha) \equiv - \log \pi(\theta, \alpha).
\end{align}
